---
title: "AETHEL: Longitudinal Impact of Spatio-Temporal Environmental Exposures on Polygenic Disease Risk"
author: "Tanya Deep"
date: today
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    code-fold: true
    embed-resources: true
    fig-width: 9
    fig-height: 6
execute:
  warning: false
  message: false
---

## Abstract

This report details the statistical methodology and findings of the AETHEL framework. By synthesising a cohort of 1,000 patients distributed across a metadata registry of 100 European geographic hubs, we evaluated the proportional hazards of respiratory events against engineered features, including cumulative ambient PM₂.₅ exposure and individualised polygenic risk scores.

---

## Methodology

### 1. Data Engineering (Python)

Patient-specific environmental exposures were calculated by aligning temporal meteorological data (simulated across 100 European healthcare hubs) with individual right-censored observation windows. Geographic scattering (±0.02°) was applied to prevent overlapping coordinate mapping in the downstream spatial analysis.

### 2. Survival Analysis (R)

We utilised a standard Cox Proportional Hazards model to establish baseline inference, defined by the hazard function:

$$
h(t, X) = h_0(t) \exp\left(\sum_{i=1}^{p} \beta_i X_i\right)
$$

Where $h_0(t)$ is the baseline hazard and $X_i$ represents our clinical, genetic, and environmental covariates.

To capture non-linear interactions and extract feature importance, we subsequently trained a Random Survival Forest (RSF) ensemble.

---

## Statistical Results

The following R code chunk loads the configuration and executes the primary biostatistical inference.

```{r}
#| label: load-config
library(yaml)
library(survival)
library(randomForestSRC)

config      <- yaml::read_yaml("../configs/default.yaml")
r_seed      <- config$seeds$r
covariates  <- config$features$survival_covariates

cohort_data <- read.csv("../data/processed/analytical_cohort.csv")
cat(sprintf("Cohort loaded: %d patients\n", nrow(cohort_data)))
```

### Cox Proportional Hazards

```{r}
#| label: cox-model

surv_obj  <- Surv(time = cohort_data$months_observed, event = cohort_data$event_occurred)
cox_formula <- as.formula(paste("surv_obj ~", paste(covariates, collapse = " + ")))
cox_model <- coxph(cox_formula, data = cohort_data)

cat("Cox Proportional Hazards Coefficients:\n")
print(summary(cox_model)$coefficients)
```

### Proportional Hazards Assumption

```{r}
#| label: ph-test

ph_test <- cox.zph(cox_model)
cat("Schoenfeld Residuals Test (Proportionality Check):\n")
print(ph_test)
```

### Explainable AI — Variable Importance

Using the `randomForestSRC` implementation (500 trees, seeded deterministically), we isolated the primary drivers of hazard risk.

```{r}
#| label: rsf-model
#| cache: true

set.seed(r_seed)
rsf_formula <- as.formula(paste(
  "Surv(months_observed, event_occurred) ~",
  paste(covariates, collapse = " + ")
))

rsf_model <- rfsrc(
  rsf_formula,
  data       = cohort_data,
  ntree      = config$model_params$rsf_ntree,
  importance = TRUE,
  seed       = r_seed
)

cat(sprintf("OOB Error Rate: %.4f\n", rsf_model$err.rate[config$model_params$rsf_ntree]))
```

```{r}
#| label: vimp-results

vimp_raw    <- vimp(rsf_model)$importance
sorted_vimp <- sort(vimp_raw, decreasing = TRUE)

cat("Ranked Variable Importance (VIMP):\n")
print(sorted_vimp)
```

---

## Conclusion

The AETHEL framework successfully bridges high-throughput spatio-temporal data engineering with rigorous, reproducible biostatistical modelling. By scaling the pipeline to 100 distinct geographic hubs, AETHEL provides a robust and scalable foundation for precision medicine and population health intelligence. Long-term PM₂.₅ exposure and baseline genomic risk emerge as the dominant predictive features across the European cohort.
